---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

## Loading and preprocessing the data

```{r options}
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r read data, results = "hide"}
activityData<-read.csv("activity.csv", header = T)
```

Once the data is read, there are a number of questions we must ask using it. 

```{r investigate data}
sapply(activityData, class)
```

One thing we are gonna need for sure is to manipulate the interval variable, so we can work with the dates on a proper way. First we convert the variable to character, then we get attach some zeros to the ones that are less than 4 characters long. This way, we make sure we can later parse them in the correct format (hours:minutes). 

```{r convert intervals}
activityData$interval<-as.character(activityData$interval)

for (i in 1:length(activityData$interval)) {
        while (nchar(activityData$interval[i])<4) {
                activityData$interval[i]<-paste("0", 
                                               activityData$interval[i], 
                                               sep = ""); 
                if(nchar(activityData$interval[i])==4) break}
}

sum(nchar(activityData$interval)!=4)
```



Now that we have reasonably clean data, we can start investigating.

```{r load libraries, results = "hide"}
library(ggplot2)
library(dplyr)
```

## What is mean total number of steps taken per day?

``` {r by date}
stepsbyday<-activityData %>% 
        group_by(date) %>% 
        summarise (total = sum(steps), 
                   mean = mean(steps, na.rm = T), 
                   median = (median(steps, na.rm = T)))
options(scipen = 6)
totalmean<-mean(stepsbyday$total, na.rm = T)
totalmedian<-median(stepsbyday$total, na.rm = T)

```

Mean steps taken by day is `r totalmean`.

Median steps taken by day is `r totalmedian`.

```{r by day histogram}
bydayhist<-qplot(stepsbyday$total)
bydayhist + geom_histogram()
```


## What is the average daily activity pattern?

```{r by interval}
intervalsteps<- activityData %>%
        group_by(interval) %>%
        summarise(average = mean(steps, na.rm =T))

#change format to something readable

intervalsteps$interval<-as.POSIXlt(intervalsteps$interval, format = "%H%M")
```

```{r steps plot}
q2<-ggplot(intervalsteps, aes(x = interval, y = average))
library(scales)
q2+geom_line() + scale_x_datetime(labels = date_format("%H:%M"))
```


## Imputing missing values

A quick check for getting to know how the na values are distributed.

```{r na table}
table(with(activityData, tapply(steps, date, FUN = function(x) sum(is.na(x)))))
```

This shows clearly that there are just two types of days: the ones on which every interval was monitored and the ones on which no interval was monitored. Hence, we can not impute the missing values using the mean/median for the day, but instead we must use the center value of the 5-minute interval.

A quick check on the differences between mean and median for each interval reveals that, in some of the intervals, differences are quite high between the 2 values:
```{r check differences}
medians<-with(activityData, tapply(steps, interval, FUN = function(x) median(x, na.rm = T)))
means<-with(activityData, tapply(steps, interval, FUN = function(x) mean(x, na.rm = T)))
differences<-means-medians
plot(differences, type = "l")
```

Considering that the max value in the means is `r max(means)`, a maximum difference of `r max(differences)` between the mean and the median seems to imply that there is a (strong) skew on the data. When data is skewed, median is usually a better indicator of the center of the distribution, since it is more robust to outliers, so we decide to impute based on that. With the data that we have, we can't do better.

```{r impute}
#copy of the dataset
activityImpute<-activityData
#compute medians for each interval
medians<-activityData %>%
        group_by(interval) %>%
        summarise(median = median (steps, na.rm = T))
#get a vector of rows that have na values
nas<-which(is.na(activityImpute$steps))
#impute based on that
for (i in 1:length(nas)){
        activityImpute[nas[i], "steps"]<- 
                medians$median[medians$interval == activityImpute[nas[i], "interval"]]
}
```

One obvious consequence of imputation using the median in this case is that, since it is (almost always) lower than the mean, general values for the distribution should have "dropped". Let's check with an histogram of steps by day (the same as the first question) but with the new data.

```{r imputed data hist}
actImputedDay<-activityImpute %>%
        group_by(date) %>%
        summarise (total = sum(steps))
meanImp<-mean(actImputedDay$total)
medianImp<-median(actImputedDay$total)
```

The mean of the new dataset with the imputed data is `r meanImp` while the mean of the original dataset before the imputation was `r totalmean`.

The median of the imputed data is `r medianImp` while the median of the original dataset before the imputation was `r totalmedian`.

Now the histogram:

```{r imputed histogram}
impdayhist<-qplot(actImputedDay$total)
impdayhist + geom_histogram()
```


## Are there differences in activity patterns between weekdays and weekends?

First we use the wday function (in the lubridate package) to classify the dates in weekdays and weekends.
It is worth reminding that wday(date) returns an integer with values in the range 1:7, starting on sunday.

Then we group the dataset by two variables, the weekday and the 5-minute interval, and get a mean for each possible combination of the two.


``` {r make weekday}
library(lubridate)
activityImpute$weekday<-ifelse(wday(activityImpute$date)>1 & #1 = sunday
                               wday(activityImpute$date)<7,  #7 = saturday
                               "weekday", 
                               "weekend")

wdaydata<-activityImpute %>%
        group_by(interval, weekday) %>%
        summarise(mean = mean(steps))

wdaydata$interval<-as.POSIXlt(wdaydata$interval, format = "%H%M")

wdaydata$weekday<-as.factor(wdaydata$weekday)
```

Then we pass those values to ggplot (the interval on the x axis, the computed mean on the y axis), using the weekday factor to facet the plot. It seems that there are different patterns for weekdays and weekends, and while, in the former, the steps seem to concentrate around some time ranges, in the latter we can see a much more regular distribution of steps taken through the day.

```{r weekday plot}
graph<-ggplot(wdaydata, aes(x = interval, y = mean))
graph + geom_line() + facet_grid(weekday~.) + scale_x_datetime(labels = date_format("%H:%M"))
```
